### 2018-12-05 chapters/2-methods.tex
fixes; readability; restructuring

In this study, the gray-level co-occurrence matrix, the local binary patterns,
and the histogram of oriented gradients can be assigned into the statistical
category, and the Sobel operator into the structural one. The Hu and Zernike
moments belong to the model-based group, while the Gabor filter and the Haar
wavelet are transform methods. The selection of texture extraction methods used
in this study was based mainly on prior experiences in existing MRI literature
\citep{Castellano2004, Lemaitre2015} and the availability of applicable
open-source software components. All of these texture descriptor methods, except
the Hu and Zernike moments, have been previously used for CAD of PCa, although
implementation details are very scarce.

@@

@@

Seven window configurations were used for DWI and nine for T₂w and T₂
data. These square-shaped windows had evenly spaced voxel side lengths of 3, 5,
…, 15 for DWI and 3, 7, …, 35 for the higher-resolution T₂w and T₂. These
lengths correspond to 3.3--17 mm for DWI and 1.4--16 mm for T₂w and T₂,
maintaining similar physical cover among differing resolutions.

For each lesion, the window was placed on all possible locations along the
transverse planes so that it still stayed completely within the lesion area. In
cases of the window not fitting completely inside lesion area, the window
locations with maximum lesion area were used. Extracted texture features were
then averaged over all slices to be used as lesion-wise features. The use of
different window sizes and parameter combinations resulted in 1281 features per
DWI image type (ADCₘ, ADCₖ, K) and 1631 features per other image type (T₂w,
T₂), totaling 7105 features all five image types combined. The free software
libraries Scikit-image \citep{VanderWalt2014} and Mahotas \citep{Coelho2013}
were utilized in the implementation of the process.

@@

In this study, the six GLCM features implemented by the Scikit-image software
library \citep{VanderWalt2014} were extracted. These features, namely contrast,
dissimilarity, homogeneity, energy, correlation, and angular second moment
(i.e.\ uniformity), are among the ones most commonly used in previous studies
\citep{Clausi2002Rapid}. They are generally considered effective texture
discriminators, and maintain invariance regarding scale and shift
\citep{Clausi2002Analysis}. In order to gain orientation invariance, the results
were averaged over four bidirectional axes, and mean range over the orientations
was added \citep{Haralick1973}. Since this method requires a discrete source
image, the images needed prior normalization depending on source type. All
images were quantized by uniform scaling to 32 gray levels based on previous
studies \citep{Clausi2002Analysis, Clausi2002Rapid, Albregtsen2008}. The image
type specific source intensity ranges for scaling were manually defined by
observing prostate volume histograms. In total 324--420 GLCM related features
with sliding window approach were extracted per image type (T₂w, ADCₘ, ADCₖ, K,
T₂).

@@

The Gabor function \citep{Gabor1946} is a Gaussian modulated by a sinusoidal
wave. Gabor filter banks can be used for texture characterization, as each
filter, shaped by its parameters, responds to specific local spatial frequency
properties of the image \citep{Turner1986}. Gabor filters are sensitive to edges
in the image, so given that different lesions contain different regions, the
detected edges between the regions could yield different responses. In this
study, the texture extraction scheme described by \citet{Tuceryan1998} was
applied, which uses the sliding window on the Gabor-filtered images.

Based on experimentation and data visualization, the filter bank included the
combinations of five different frequencies for the sinusoidal
($f=0.1,0.2,…,0.5$ per voxel), three sizes for a circular Gaussian
envelope ($\sigma=1,2,3$), and four bidirectional orientations. A number of
derived features have been suggested \citep{Clausi2000, Hammouda2000,
Grigorescu2002}. Here, the mean of the real part, variance of the real part,
mean of the absolute of the real part, and mean of the magnitude were extracted
as features. Various ways to achieve orientation invariance have been proposed
\citep{Arivazhagan2006, Han2007, Chu2009, Rahman2011}, although not all are
equally suitable for texture classification. In this study, the simple method of
summing filtered images over orientations was used \citep{Han2007}, yielding
420--540 features per image type.

@@

Image moments are weighted averages that describe the distribution of intensity
within image. A few variations are widely used as object shape descriptors, but
they have also been applied for texture analysis \citep{Tuceryan1994}. In this
study, the absolute values of the seven Hu moments \citep{Hu1962}, and the
magnitude values of the complex Zernike moments \citep{Teague1980} up to the 8th
degree were calculated for each window. The Hu moments are invariant regarding
translation, scaling, rotation, and reflection \citep{Theodoridis2003}. The
magnitude of the Zernike moments are rotation invariant, robust to noise, and,
due to their orthogonality, have minimal redundancy \citep{Tahmasbi2011,
Amayeh2005}. In total, 49--63 Hu moments and 175--225 Zernike moments were
extracted from each image type.

@@

The Sobel operator \citep{Sobel1990} is a simple convolution filter that
emphasizes edges in image. The shape of the kernel window is 3×3 by definition,
so no other window sizes were used. Instead, the mean of the whole lesion-wide
texture map was used, both with the lesion edge voxels included and excluded,
resulting in 2 features.

In addition, 18 first-order statistical features were calculated for the whole
lesion. First-order statistics imply that only the probabilistic distribution of
intensity values is of concern, without observing their spatial relations. These
statistical features were mean, standard deviation, range, minimum, maximum, the
quartiles, the deciles, kurtosis, and skewness. Existing literature typically
utilizes averages and percentiles \citep{Shaish2016}.

>>>>

In this study, the gray-level co-occurrence matrix, the local binary patterns,
and the histogram of oriented gradients can be assigned into the statistical
category, and the Sobel operator into the structural one. The Hu and Zernike
moments belong to the model-based group, while the Gabor filter and the Haar
wavelet are transform methods.

The selection of texture extraction methods used in this study was based mainly
on prior experiences in existing MRI literature \citep{Castellano2004,
Lemaitre2015} and the availability of applicable open-source software
components. All of these texture descriptor methods, except the Hu and Zernike
moments, have been previously used for CAD of PCa. However, detailed information
on their implementation and parameter selection are usually very scarce. This
issue was addressed in this study by providing complete information on the
parameters, and by using a wide array of them.

@@

\subsubsection{General implementation details}

@@

Seven window configurations were used for DWI and nine for T₂w and T₂
data. These square-shaped windows had evenly spaced voxel side lengths of 3, 5,
…, 15 for DWI and 3, 7, …, 35 for the higher-resolution T₂w and T₂. These
lengths correspond to 3.3--17 mm for DWI and 1.4--16 mm for T₂w and T₂,
maintaining similar physical cover over different resolutions.

For each lesion, the window was placed on all possible locations along the
transverse planes so that it still stayed completely within the lesion area. In
cases of the window not fitting completely inside lesion area, the window
locations with maximum lesion area were used. Extracted texture feature maps
were then averaged over all slices to be used as lesion-wise median features.
The use of different window sizes and parameter combinations resulted in 1281
features per DWI image type (ADCₘ, ADCₖ, K) and 1631 features per other image
type (T₂w, T₂), totaling 7105 features all five image types combined. The free
software libraries Scikit-image \citep{VanderWalt2014} and Mahotas
\citep{Coelho2013} were utilized in the implementation of the process.

\subsubsection{Method-specific implementation details}

@@

In this study, the six GLCM features implemented by the Scikit-image software
library \citep{VanderWalt2014} were extracted. These features, namely contrast,
dissimilarity, homogeneity, energy, correlation, and angular second moment
(i.e.\ uniformity), are among the ones most commonly used in previous studies
\citep{Clausi2002Rapid}. They are generally considered effective texture
discriminators, and maintain invariance regarding scale and shift
\citep{Clausi2002Analysis}. In order to gain orientation invariance, the results
were averaged over four bidirectional axes, and mean range over the orientations
was added \citep{Haralick1973}.

Since GLCM requires a discrete source image, the images needed prior
normalization depending on source type. All images were quantized by uniform
scaling to 32 gray levels, based on previous studies \citep{Clausi2002Analysis,
Clausi2002Rapid, Albregtsen2008}. The image type specific source intensity
ranges for scaling were manually defined by observing prostate volume
histograms. In total 324--420 GLCM related features with sliding window approach
were extracted per image type (T₂w, ADCₘ, ADCₖ, K, T₂).

@@

The Gabor function \citep{Gabor1946} is a Gaussian modulated by a sinusoidal
wave. Gabor filter banks can be used for texture characterization, as each
filter, shaped by its parameters, responds to specific local spatial frequency
properties of the image \citep{Turner1986}. Gabor filters are sensitive to edges
in the image, so given that different lesions contain different regions, the
detected edges between the regions could yield different responses. In this
study, the texture extraction scheme described by \citet{Tuceryan1998} was
applied, which uses the sliding window on the Gabor-filtered complex images.

Based on experimentation and data visualization, the filter bank included the
combinations of five different frequencies for the sinusoidal
($f=0.1,0.2,…,0.5$ per voxel), three sizes for a circular Gaussian
envelope ($\sigma=1,2,3$), and four bidirectional orientations. A number of
derived features have been suggested \citep{Clausi2000, Hammouda2000,
Grigorescu2002}. Here, the extracted features included mean of the real part,
variance of the real part, mean of the absolute of the real part, and mean of
the magnitude. Various ways to achieve orientation invariance have been proposed
\citep{Arivazhagan2006, Han2007, Chu2009, Rahman2011}, although not all are
equally suitable for texture classification. In this study, the simple method of
summing filtered images over orientations was used \citep{Han2007}, yielding
420--540 features per image type.

@@

Image moments are weighted averages that describe the distribution of intensity
within image. A few variations are widely used as object shape descriptors, but
they have also been applied to texture analysis \citep{Tuceryan1994}. In this
study, logarithms of the absolute values of the seven Hu moments \citep{Hu1962},
and the magnitude values of the complex Zernike moments \citep{Teague1980} up to
the 8th degree were calculated for each window. The Hu moments are invariant
regarding translation, scaling, rotation, and reflection
\citep{Theodoridis2003}. The Zernike moment magnitudes are rotation invariant,
robust to noise, and, due to their orthogonality, have minimal redundancy
\citep{Tahmasbi2011, Amayeh2005}. In total, 49--63 Hu moments and 175--225
Zernike moments were extracted from each image type.

@@

The Sobel operator \citep{Sobel1990} is a simple convolution filter that
emphasizes edges in image. The shape of the kernel window is 3×3 by definition,
so no other window sizes were used. Instead, the median of the whole lesion-wide
texture map was used, both with the lesion edge voxels included and excluded,
resulting in 2 features.

In addition, first-order statistical features were calculated over the whole
lesion. First-order statistics observe only the probabilistic distribution of
intensity values, ignoring their spatial relations. Existing literature
typically utilizes averages and some of the percentiles \citep{Shaish2016}.
Here, 18 features were included, namely mean, standard deviation, range,
minimum, maximum, quartiles, deciles, kurtosis, and skewness.
